---
title: "Interpreting time in buzzr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting time in buzzr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(buzzr)
```

buzzr works with two different representations of time:

- **File-time** (`start_filetime`): seconds elapsed since the start of the audio file. This is what buzzdetect writes directly — it knows nothing about when you pressed record.
- **Date-time** (`start_datetime`): an absolute timestamp, like `2023-08-09 06:00:00`. To get this, buzzr needs to know when the recording started, which it reads from the file name.

This vignette explains how buzzr converts file names into date-times, how to handle mixed recorder types, and how to work with the resulting timestamps for plotting and modelling.

---

## Reading a timestamp from a file name

Most field recorders embed the recording start time in the file name.
buzzr reads this using the same POSIX format codes as base R's `strptime()`.
Common examples:

| File name pattern | Format string |
|---|---|
| `230809_0000` (YYMMDD_HHMM) | `'%y%m%d_%H%M'` |
| `20230809_000000` (YYYYMMDD_HHMMSS) | `'%Y%m%d_%H%M%S'` |
| `2023-08-09_00-00-00` | `'%Y-%m-%d_%H-%M-%S'` |

The Five Flowers dataset uses the `YYMMDD_HHMM` pattern.
Let's parse a single file:

```{r single-file}
path <- system.file(
  'extdata/five_flowers/soybean/9/230809_0000_buzzdetect.csv',
  package = 'buzzr'
)

# Without posix_formats: time stays as file-time (seconds from file start)
head(read_results(path), 3)

# With posix_formats: file-time is converted to an absolute date-time
head(read_results(path, posix_formats = '%y%m%d_%H%M', tz = 'America/New_York'), 3)
```

The `_buzzdetect` suffix and file extension are stripped before matching,
so your format string only needs to describe the timestamp part of the name.

You can also call `file_start_time()` directly if you just want the recording's
start time without reading the full file:

```{r file-start-time}
file_start_time(path, posix_formats = '%y%m%d_%H%M', tz = 'America/New_York')
```

It works on a vector of paths too:

```{r file-start-time-vector}
dir <- system.file('extdata/five_flowers', package = 'buzzr')
paths <- list.files(dir, pattern = '_buzzdetect', recursive = TRUE, full.names = TRUE)

file_start_time(paths, posix_formats = '%y%m%d_%H%M', tz = 'America/New_York')
```

---

## Time zones

Always pass a `tz` argument.
If you omit it, R uses your system time zone, which will silently produce wrong
results if you analyse data collected in a different zone.

```{r tz-comparison}
# Same file, two different time zones — note the hour difference
file_start_time(path, posix_formats = '%y%m%d_%H%M', tz = 'America/New_York')
file_start_time(path, posix_formats = '%y%m%d_%H%M', tz = 'America/Chicago')
```

Use `OlsonNames()` to browse valid time zone strings.

---

## Keeping both time columns

By default, `start_filetime` is dropped once `start_datetime` is added.
Set `drop_filetime = FALSE` to keep both:

```{r drop-filetime}
head(
  read_results(
    path,
    posix_formats = '%y%m%d_%H%M',
    tz = 'America/New_York',
    drop_filetime = FALSE
  ),
  3
)
```

File-time is useful if you want to extract audio clips later
(e.g. using FFmpeg to cut from a specific second in the source file).

---

## Mixed recorder types: multiple format strings

If your experiment used more than one recorder model, the file names may follow
different timestamp conventions. Supply all formats as a character vector:

```{r multiple-formats-setup}
# Imagine two recorder types in the same experiment:
#   AudioMoth:  YYYYMMDD_HHMMSS  e.g. 20230809_060000_buzzdetect.csv
#   SongMeter:  YYMMDD_HHMM      e.g. 230809_0600_buzzdetect.csv

paths_mixed <- c(
  'site_a/audiomoth/20230809_060000_buzzdetect.csv',  # AudioMoth
  'site_b/songmeter/230809_0600_buzzdetect.csv'       # SongMeter
)

formats_mixed <- c('%Y%m%d_%H%M%S', '%y%m%d_%H%M')
```

When each file matches exactly one format, everything resolves cleanly:

```{r multiple-formats-clean}
file_start_time(paths_mixed, posix_formats = formats_mixed, tz = 'America/New_York')
```

### When a file matches two formats with different results

This is where things get tricky.
A SongMeter file like `230809_0600` will match `%y%m%d_%H%M` (→ 2023) but it
can also match `%Y%m%d_%H%M%S` if interpreted as year 0230 (an implausible
date, but R won't reject it).
More dangerously, `20230809_060000` matches `%Y%m%d_%H%M%S` (→ 2023-08-09
06:00:00) but also `%y%m%d_%H%M` (→ year 2020, completely wrong).

By default (`first_match = FALSE`), buzzr returns `NA` and warns you when
formats conflict:

```{r first-match-false}
# Construct a path that matches both formats ambiguously
ambiguous_path <- 'site/recorder/20230809_060000_buzzdetect.csv'

file_start_time(
  ambiguous_path,
  posix_formats = c('%Y%m%d_%H%M%S', '%y%m%d_%H%M'),
  tz = 'America/New_York',
  first_match = FALSE
)
```

If you know which format should take priority, set `first_match = TRUE` to
accept the first matching format and move on:

```{r first-match-true}
# With '%Y%m%d_%H%M%S' listed first, the four-digit-year parse wins
file_start_time(
  ambiguous_path,
  posix_formats = c('%Y%m%d_%H%M%S', '%y%m%d_%H%M'),
  tz = 'America/New_York',
  first_match = TRUE
)

# Swap the order — now the two-digit-year parse wins (wrong!)
file_start_time(
  ambiguous_path,
  posix_formats = c('%y%m%d_%H%M', '%Y%m%d_%H%M%S'),
  tz = 'America/New_York',
  first_match = TRUE
)
```

The order of `posix_formats` matters when `first_match = TRUE`.
Put the most specific (or most trusted) format first.

### When a file matches two formats with the *same* result

If two formats resolve to identical timestamps, buzzr silently deduplicates
them — no warning, no NA:

```{r same-result}
# These two formats both parse '230809_0600' as 2023-08-09 06:00:00
# (one by coincidence of the numeric overlap)
file_start_time(
  'site/recorder/230809_0600_buzzdetect.csv',
  posix_formats = c('%y%m%d_%H%M', '%y%m%d_%H%M'),  # duplicate format
  tz = 'America/New_York'
)
```

### If no format matches

buzzr returns `NA` with a warning, so downstream code doesn't fail silently:

```{r no-match}
file_start_time(
  'site/recorder/no_timestamp_here_buzzdetect.csv',
  posix_formats = '%y%m%d_%H%M',
  tz = 'America/New_York'
)
```

---

## Plotting time of day: `commontime()`

When recordings span multiple calendar days you can't overlay them directly on
a datetime axis — the dates differ even if the time of day is the same.
`commontime()` solves this by setting every date to the same arbitrary day
(2000-01-01) while preserving the time of day:

```{r commontime}
dir <- system.file('extdata/five_flowers', package = 'buzzr')

binned <- bin_directory(
  dir_results   = dir,
  thresholds    = c(ins_buzz = -1.2),
  posix_formats = '%y%m%d_%H%M',
  tz            = 'America/New_York',
  dir_nesting   = c('flower', 'recorder'),
  binwidth       = 20,
  calculate_rate = TRUE
)

# Dates vary across the five recordings; commontime collapses them
range(binned$bin_datetime)

binned$time_of_day <- commontime(binned$bin_datetime, tz = 'America/New_York')
range(binned$time_of_day)
```

Now all five flowers share the same x axis, regardless of which day they were
recorded:

```{r commontime-plot, fig.width=7, fig.height=3.5}
library(ggplot2)

ggplot(binned, aes(x = time_of_day, y = detectionrate_ins_buzz, color = flower)) +
  geom_line() +
  scale_x_datetime(labels = label_hour(tz = 'America/New_York')) +
  labs(x = 'Time of day', y = 'Detection\nrate', color = 'Flower') +
  theme_buzzr()
```

Always pass the **same `tz`** to `commontime()` and `label_hour()`.
If they differ, the axis labels won't match the data.

---

## Numeric time of day: `time_of_day()`

For statistical modelling — circular regression, GAMs, peak-time estimates —
it's more convenient to have time as a number.
`time_of_day()` returns either a proportion of the 24-hour day or a decimal hour:

```{r time-of-day}
times <- as.POSIXct(
  c('2023-08-09 00:00:00', '2023-08-09 06:00:00',
    '2023-08-09 12:00:00', '2023-08-09 18:30:00'),
  tz = 'America/New_York'
)

# Proportion: midnight = 0, noon = 0.5, next midnight = 1
time_of_day(times)

# Decimal hours: midnight = 0, 6 AM = 6, 6:30 PM = 18.5
time_of_day(times, time_format = 'hour')
```

Unlike `commontime()`, this strips the date entirely and returns a plain number,
which is easier to pass to modelling functions:

```{r time-of-day-model}
binned$hour <- time_of_day(binned$bin_datetime, time_format = 'hour')

# e.g. find the hour of peak detections for each flower
binned[, .SD[which.max(detectionrate_ins_buzz)][, .(peak_hour = hour)],
       by = flower]
```
